---
title: "yolo: You Only Lookup One (small part of an otherwise large matrix)"
author: "Caleb Lareau"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{yolo: You Only Lookup One (small part of an otherwise large matrix).}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## About
The R package `yolo` is designed to subset large data based on column and row
attributes based on the familiar `RangedSummarizedExperiment` (rse) structure without
holding the matrix of values in memory. To achieve this, an `rseHandle` S4 object
is defined that inherits the `RangedSummarizedExperiment` class with the addition
of two other slots that map the current object's row and column indicies to the 
original indicies in the file(s). The `getvalues` command can then evaluate an
`rseHandle` object and pull the data from the hard disk into memory. While adding and
subsetting an `rseHandle` object is endomorphic (i.e. returns an `rseHandle`),
the output of the `getvalues` is a `RangedSummarizedExperiment` object. 

## Libraries
```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(GenomicRanges)
library(SummarizedExperiment)
library(yolo)
library(RSQLite)
library(rhdf5)
```

## Build Data--SQLite
In the current implementation of `yolo`, we support storing data in HDF5
and sqlite file formats. Below is one simple implementation of converting
a .csv file that is in a sparse matrix format into a `.sqlite` object. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
f1name <- "d1.sqlite"

db <- dbConnect(SQLite(), dbname=f1name)
ft <- list(row="INTEGER", column="INTEGER", value="INTEGER")
df1 <- read.table(system.file("extdata", "dat1.csv", package = "yolo"), sep = "," , header = TRUE)
head(df1)
RSQLite::dbWriteTable(conn=db, name="data", value=df1, field.types=ft, overwrite = TRUE)
dbDisconnect(db)
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
readt <- read.table(system.file("extdata", "dat1_row.bed", package = "yolo"))
rowData1 <- GRanges(setNames(readt,  c("chr", "start", "stop")))
colData1 <- read.table(system.file("extdata", "dat1_col.txt", package = "yolo"))
```

Now that the column and row data annotations have been imported into our system,
we can build our `rseHandle` object using the following function below. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
d1 <- rseHandleMake(rowData1, colData1, lookupFileName = f1name)
d1
```
The `rseHandleMake` function necessarily takes a `GRanges` of the rowData,
an object that can be coerced into a DataFrame for the colData, and a valid
file name that contains the values of the matrices on the backend. When the
constructor function is called, other checks will determine the validity of 
the construction to ensure that the specified objects will play nicely 
together. In other words, the constructor checks to make sure the dimensions
of the rowData and colData represent the dimensions in the backend file. <br><br>

Three other parameters can be specified in the `rseHandleMake` function. Namely,
the `lookupTableName` specifies the index/table of the backend values. By default,
the constructor assumes "data" as we specified in the `dbWriteTable` command earlier
in the vignette. Another import parameter is the `lookupFileType`, which can be
specified as either "sparse" (by default) or "normal". For a sparse matrix,
we assume two columns labled "row" and "column" in addition to a third that has
the specific values. (See the `ft` variable in the constructor above). For a "normal"
matrix, the lookup simply indexes off of row and column positions, so that the
names are not relevant for that operation. Finally, the `lookupFileFormat` can
be either "HDF5" or "SQLite." The call to the `rseHandleMake` function above
utilized all default values--

(`lookupTableName` = "data", `lookupFileType` = "sparse", `lookupFileFormat` = "sqlite")

## Build Data-- HDF5
Another implementation uses HDF5. Currently, `yolo` only supports the normal
matrix implementation (sparse matricies are not supported). This is because 
the author couldn't find a way to filter to rows based on values. This package
supports putting multiple tables in either an HDF5 or SQLite file, and the 
implementation would look similar to the following. 

```{r, echo = TRUE, message = FALSE, warning = FALSE}
f2name <- "dat.hdf5"
h5createFile(f2name)

# Read and Reshape 3 data objects to a normal matrix
df1 <- read.table(system.file("extdata", "dat1.csv", package = "yolo"), sep = "," , header = TRUE)
dat1m <- reshape2::acast(df1, row ~ column, fill = 0)
df2 <- read.table(system.file("extdata", "dat2.csv", package = "yolo"), sep = "," , header = TRUE)
dat2m <- reshape2::acast(df2, row ~ column, fill = 0)
df3 <- read.table(system.file("extdata", "dat3.csv", package = "yolo"), sep = "," , header = TRUE)
dat3m <- reshape2::acast(df3, row ~ column, fill = 0)

# Write to file
h5write(dat1m, "dat.hdf5","dat1")
h5write(dat2m, "dat.hdf5","dat2")
h5write(dat3m, "dat.hdf5","dat3")

h5ls("dat.hdf5")
```

To create an `rseHandle` for the first dataset--

```{r, echo=TRUE, message=FALSE, warning=FALSE}
d1h <- rseHandleMake(rowData1, colData1, lookupFileName = f2name, lookupTableName = "dat1",
                     lookupFileFormat = "HDF5", lookupFileType = "normal")
d1h
```

We'll also create an `rseHandle` object for the third data object referencing 
the same HDF5 file but different colData. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
colData3 <- read.table(system.file("extdata", "dat3_col.txt", package = "yolo"))
d3h <- rseHandleMake(rowData1, colData3, lookupFileName = f2name, lookupTableName = "dat3",
                     lookupFileFormat = "HDF5", lookupFileType = "normal")
```

## Addition
Users can add multiple `rseHandle` objects together as long as
two condtions are valid--

1) The `rowRanges/rowData` are the same
2) The names in `colData` are the same

Even though `d1` pulls from a sparse sqlite file and 
`d3h` pulls from a normal HDF5 file, these two handles
can be joined together because these two criteria are met. 
Notice that the resulting object has 35 samples. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
d13 <- d1 + d3h
d13
```

This feature allows for samples from different experiments to
be joined together at a high level again without reading any value data
into memory aside from the column and row meta data. 

## Subsetting
Users can subset using the `[` and `subsetByOverlaps` calls that they
are accustomed to in a standard `RangedSummarizedExperiment`. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
dss1 <- d13[,d13@colData$group == "group4"]
chr1reg <- GRanges(seqnames=c("chr1"),ranges=IRanges(start=c(3338300),end=c(3422000)))
dss2 <- subsetByOverlaps(dss1, chr1reg)
d_small <- dss2[c(2,3,6,7,10), c(2,6,7,3)]
d_small
```

## Get values
Through this process of adding and subsetting, we've jumbled up our samples. 
Not to worry! Using the `getvalues` function, the representation of our 
matrix is data will be preserved through keeping track of the indices
of our files, rows, and columns. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
rse_small <- getvalues(d_small)
assay(rse_small, 1)
```

## Cleanup
Without any use for our files on disk, we can tidy up and remove
them. 
```{r, echo=TRUE, message=FALSE, warning=FALSE}
file.remove(f1name)
file.remove(f2name)
```

## Session info
```{r sessInfo}
sessionInfo()
```
