---
title: "yolo: You Only Lookup One (small part of an otherwise large matrix)"
author: "Caleb Lareau"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{yolo: You Only Lookup One (small part of an otherwise large matrix).}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## About
The R package `yolo` is designed to subset large data based on column and row
attributes based on the familiar `RangedSummarizedExperiment` structure without
holding the matrix of values in memory. To achieve this, an `rseHandle` S4 object
is defined that inherits the `RangedSummarizedExperiment` class with the addition
of two other slots that map the current object's row and column indicies to the 
original indicies in the file. The `getvalues` command can then evaluate an
`rseHandle` object and pull the data from the hard disk into memory. While 
subsetting an `rseHandle` object is endomorphic (i.e. returns an `rseHandle`),
the output of the `getvalues` is a `RangedSummarizedExperiment` object. 

## Libraries
```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(GenomicRanges)
library(yolo)
library(RSQLite)
```

## Build Data
In the current implementation of `yolo`, we support storing data in HDF5
and sqlite file formats. Below is one simple implementation of converting
a .csv file that is in a sparse matrix format into a `.sqlite` object. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
f1name <- "d1.sqlite"

db <- dbConnect(SQLite(), dbname=f1name)
ft <- list(row="INTEGER", column="INTEGER", value="INTEGER")
df3 <- read.table(system.file("extdata", "dat1.csv", package = "yolo"), sep = "," , header = TRUE)
head(df3)
RSQLite::dbWriteTable(conn=db, name="data", value=df3, field.types=ft, overwrite = TRUE)
dbDisconnect(db)
```


```{r, echo=TRUE, message=FALSE, warning=FALSE}
readt <- read.table(system.file("extdata", "dat1_row.bed", package = "yolo"))
rowData <- GRanges(setNames(readt,  c("chr", "start", "stop")))
colData <- read.table(system.file("extdata", "dat1_col.txt", package = "yolo"))
```

Now that the column and row data annotations have been imported into our system,
we can build our `rseHandle` object using the following function below. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
d1 <- rseHandleMake(rowData, colData, lookupFileName = f1name)
d1
```
The `rseHandleMake` function necessarily takes a `GRanges` of the rowData,
an object that can be coerced into a DataFrame for the colData, and a valid
file name that contains the values of the matrices on the backend. When the
constructor function is called, other checks will determine the validity of 
the construction to ensure that the specified objects will play nicely 
together. In other words, the constructor checks to make sure the dimensions
of the rowData and colData represent the dimensions in the backend file. <br><br>

Three other paraters can be specified in the `rseHandleMake` function. Namely,
the `lookupTableName` specifies the index/table of the backend values. By default,
the constructor assumes "data" as we specified in the `dbWriteTable` command earlier
in the vignette. Another import parameter is the `lookupFileType`, which can be
specified as either "sparse" (by default) or "normal". For a sparse matrix,
we assume two columns labled "row" and "column" in addition to a third that has
the specific values. (See the `ft` variable in the constructor). For a "normal"
matrix, the lookup simply indexes off of row and column positions, so that the
names are not relevant for that operation. 

## Cleanup
Without any use for our files on disk, we can tidy up and remove
them. 
```{r, echo=TRUE, message=FALSE, warning=FALSE}
file.remove(f1name)
```
## Session info
```{r sessInfo}
sessionInfo()
```
